WSClean: an implementation of a fast, generic wide-field imager

### 2014-03-17 articles/2014/Article-WSClean/wsclean.tex
Adding reference to Rau and Cornwell
Removing appendix as suggested by Tingay

For low-frequency telescopes, it can be of interest to do full-sky (i.e., horizon to horizon) imaging. In the case of MWA, the tile beam can have strong sidelobes at a distance of more than 90\degree from the pointing centre. Imaging these sidelobes might be relevant because they are scientifically of interest (e.g. when searching for transients). Full-sky imaging can also be useful for self-calibration or deconvolution. For example, self-calibration using full-sky clean components has been found to give good results in imaging the resolved FR-II radio source Fornax~A (McKinley et al., in prep.). An example full-sky MWA image is shown in Fig.~\ref{fig:full-sky-example}.

@@

Instead of imaging the whole sky, the image can be recentred from $(l,m)$ to $(\hat{l},\hat{m})=(l+\Delta l,m+\Delta m)$ by performing the substitution $(l,m)\leftarrow(\hat{l},\hat{m})$ in Eq.~\eqref{eq:wstacking}:
\begin{align}\notag
\frac{I'(\hat{l}, \hat{m})\left(w_{\max} - w_{\min}\right)}{\sqrt{1-\hat{l}^2-\hat{m}^2}} = \int\limits_{w_{\min}}^{w_{\max}} e^{2\pi i w(\sqrt{1-\hat{l}^2-\hat{m}^2}-1)} \times \\
\iint e^{2 \pi i \left(u\Delta l + v\Delta m\right)} V(u,v,w) e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
In words, recentering an image involves accounting for the position shift during the $w$-correction and final scaling, and shifting the visibilities in phase by multiplication with $e^{2 \pi i \left(u\Delta l + v\Delta m\right)}$ prior to the imaging. This essentially is a method of implementing the $w$-snapshots algorithm as described by \citet{widefield-imaging-ska-cornwell}. However, instead of creating warped images by performing the FFT over a tilted plane in $uvw$-space, our method transforms planes with constant $w$-values. The benefit of our method is that the resulting image has a circular synthesized beam, i.e., the resolution in $l$ and $m$ directions matches the intrinsic resolution of the instrument, which is desirable for cleaning. Regridding a recentred image is more straightforward compared to regridding a warped snapshot, because in the case of warped snapshots there is no analytic solution to the coordinate conversion \citep{perley-noncoplanar-arrays}. A benefit of warped snapshots is that the $w$-term errors are zero at image centre and get worse with the distance from image centre. With our approach, the effect gets worse with the distance from zenith. Our technique has a similar computational cost compared to the $w$-snapshot algorithm, although the required number of $w$-layers has a different dependency on ZA, FOV and the non-coplanarity.

@@

The prolate spheroidal wave function (PSWF) is generally considered to be the optimal windowing function for gridding \citep{fourier-kernel-selection-1991}. \textsc{casa} convolves samples with a PSWF of seven pixels total width during gridding. When using a variable kernel size, a PSWF is quite complicated and computationally expensive to calculate. \textsc{wsclean} currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and is a good approximation of the PSWF \citep{fourier-kernel-selection-1991}. After some benchmarks (see Appendix~\ref{sec:gridding-appendix}) we found that seven pixels is a good compromise between performance and accuracy, and this is the default for \textsc{wsclean}.

@@

\textsc{wsclean} is currently not able to perform multi-scale clean. Results of applying \textsc{casa}'s multi-scale clean \citep{multiscale-clean-cornwell-2008} on MWA data show that especially the Galactic plane is significantly better deconvolved using multi-scale clean, but it is very expensive. Imaging a one-minute observation takes 30 hours of computational time without $w$-projection. We plan to implement some form of multi-scale cleaning in \textsc{wsclean}. It is likely that additional optimizations need to be made to be able to do this with acceptable performance.

@@

\section*{Acknowledgments}
Offringa would like to thank O.M. Smirnov for testing \textsc{wsclean} on VLA data.
This scientific work makes use of the Murchison Radio-astronomy Observatory, operated by CSIRO. We acknowledge the Wajarri Yamatji people as the traditional owners of the Observatory site. Support for the MWA comes from the U.S. National Science Foundation (grants AST-0457585, PHY-0835713, CAREER-0847753, and AST-0908884), the Australian Research Council (LIEF grants LE0775621 and LE0882938), the U.S. Air Force Office of Scientific Research (grant FA9550-0510247), and the Centre for All-sky Astrophysics (an Australian Research Council Centre of Excellence funded by grant CE110001020). Support is also provided by the Smithsonian Astrophysical Observatory, the MIT School of Science, the Raman Research Institute, the Australian National University, and the Victoria University of Wellington (via grant MED-E1799 from the New Zealand Ministry of Economic Development and an IBM Shared University Research Grant). The Australian Federal government provides additional support via the Commonwealth Scientific and Industrial Research Organisation (CSIRO), National Collaborative Research Infrastructure Strategy, Education Investment Fund, and the Australia India Strategic Research Fund, and Astronomy Australia Limited, under contract to Curtin University. We acknowledge the iVEC Petabyte Data Store, the Initiative in Innovative Computing and the CUDA Center for Excellence sponsored by NVIDIA at Harvard University, and the International Centre for Radio Astronomy Research (ICRAR), a Joint Venture of Curtin University and The University of Western Australia, funded by the Western Australian State government. 
\appendix

\section{Optimal gridding kernel size} \label{sec:gridding-appendix}
\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/benchmark-kernelsize/kernel}
\caption{Effect of anti-aliasing kernel size on performance.}
\label{fig:timing-kernelsize}
\end{center}
\end{figure}
As discussed in \S\ref{sec:gridding}, \textsc{wsclean} uses a Kaiser-Bessel window against aliasing. The type of window function has no effect on the gridding performance, but the size of the kernel affects gridding performance quadratically. Fig.~\ref{fig:timing-kernelsize} shows that this quadratic relation becomes significant for kernel sizes $\gtrapprox 10$ pixels. Decreasing the kernel size to values below seven pixels has little effect on performance, because for small kernels the visibility-reading rate is lower than the gridding rate. We have not noticed much benefit of larger kernels except in rare cases where a bright source lies just outside the imaged FOV. Therefore, \textsc{wsclean} uses a default of seven pixels for the gridding kernel. It can be increased when necessary. Of course, different hardware might give slightly different results because of different reading and calculation performance.

>>>>

For low-frequency telescopes, it can be of interest to do full-sky (i.e., horizon to horizon) imaging. In the case of MWA, the tile beam can have strong sidelobes at a distance of more than 90\degree from the pointing centre. Imaging these sidelobes might be relevant because they are scientifically of interest (e.g. when searching for transients). Full-sky imaging can also be useful for self-calibration or deconvolution. For example, self-calibration using full-sky clean components has been found to give good results in imaging the resolved FR-II radio source Fornax~A (McKinley et al., in prep.). An example of a full-sky MWA image is shown in Fig.~\ref{fig:full-sky-example}.

@@

Instead of imaging the whole sky, the image can be recentred from $(l,m)$ to $(\hat{l},\hat{m})=(l+\Delta l,m+\Delta m)$ by performing the substitution $(l,m)\leftarrow(\hat{l},\hat{m})$ in Eq.~\eqref{eq:wstacking}:
\begin{align}\notag
\frac{I'(\hat{l}, \hat{m})\left(w_{\max} - w_{\min}\right)}{\sqrt{1-\hat{l}^2-\hat{m}^2}} = \int\limits_{w_{\min}}^{w_{\max}} e^{2\pi i w(\sqrt{1-\hat{l}^2-\hat{m}^2}-1)} \times \\
\iint e^{2 \pi i \left(u\Delta l + v\Delta m\right)} V(u,v,w) e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
In words, recentering an image involves accounting for the position shift during the $w$-correction and final scaling, and shifting the visibilities in phase by multiplication with $e^{2 \pi i \left(u\Delta l + v\Delta m\right)}$ prior to the imaging. This essentially is a method of implementing the $w$-snapshots algorithm as described by \citet{widefield-imaging-ska-cornwell}. However, instead of creating warped images by performing the FFT over a tilted plane in $uvw$-space, our method transforms planes with constant $w$-values. The benefit of our method is that the resulting image has a circular synthesized beam, i.e., the resolution in $l$ and $m$ directions matches the intrinsic resolution of the instrument, which is desirable for cleaning. Regridding a recentred image is also more straightforward compared to regridding a warped snapshot, because in the latter case there is no analytic solution to the coordinate conversion \citep{perley-noncoplanar-arrays}. A benefit of warped snapshots is that the $w$-term errors are zero at image centre and get worse with the distance from image centre. With our approach, the effect gets worse with the distance from zenith. Our technique has a similar computational cost compared to the $w$-snapshot algorithm, although the required number of $w$-layers has a different dependency on ZA, FOV and the non-coplanarity.

@@

The prolate spheroidal wave function (PSWF) is generally considered to be the optimal windowing function for gridding \citep{fourier-kernel-selection-1991}. \textsc{casa} convolves samples with a PSWF of seven pixels total width during gridding. When using a variable kernel size, a PSWF is quite complicated and computationally expensive to calculate. \textsc{wsclean} currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and is a good approximation of the PSWF \citep{fourier-kernel-selection-1991}.

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/benchmark-kernelsize/kernel}
\caption{Effect of anti-aliasing kernel size on performance.}
\label{fig:timing-kernelsize}
\end{center}
\end{figure}

The type of window function has no effect on the gridding performance, but the size of the kernel affects gridding performance quadratically. Fig.~\ref{fig:timing-kernelsize} shows that this quadratic relation becomes significant for kernel sizes $\gtrapprox 10$ pixels. Decreasing the kernel size to values below seven pixels has little effect on performance, because for small kernels the visibility-reading rate is lower than the gridding rate. We have not noticed much benefit of larger kernels except in rare cases where a bright source lies just outside the imaged FOV. Therefore, \textsc{wsclean} uses a default of seven pixels for the gridding kernel. It can be increased when necessary. Of course, different hardware might give slightly different results because of different reading and calculation performance.

@@

\textsc{wsclean} is currently not able to perform multi-scale clean. Results of applying \textsc{casa}'s multi-scale clean \citep{multiscale-clean-cornwell-2008} on MWA data show that especially the Galactic plane is significantly better deconvolved using multi-scale clean, but it is very expensive. Imaging a one-minute observation takes 30 hours of computational time without $w$-projection. We plan to implement some form of multi-scale cleaning in \textsc{wsclean}. It is likely that additional optimizations need to be made to be able to do this with acceptable performance. Combining multi-scale with wide-band deconvolution techniques is a possible further improvement \citep{rau-msmfs-2011}.

@@

\section*{Acknowledgments}
Offringa would like to thank O.M. Smirnov for testing \textsc{wsclean} on VLA data.
This scientific work makes use of the Murchison Radio-astronomy Observatory, operated by CSIRO. We acknowledge the Wajarri Yamatji people as the traditional owners of the Observatory site. Support for the MWA comes from the U.S. National Science Foundation (grants AST-0457585, PHY-0835713, CAREER-0847753, and AST-0908884), the Australian Research Council (LIEF grants LE0775621 and LE0882938), the U.S. Air Force Office of Scientific Research (grant FA9550-0510247), and the Centre for All-sky Astrophysics (an Australian Research Council Centre of Excellence funded by grant CE110001020). Support is also provided by the Smithsonian Astrophysical Observatory, the MIT School of Science, the Raman Research Institute, the Australian National University, and the Victoria University of Wellington (via grant MED-E1799 from the New Zealand Ministry of Economic Development and an IBM Shared University Research Grant). The Australian Federal government provides additional support via the Commonwealth Scientific and Industrial Research Organisation (CSIRO), National Collaborative Research Infrastructure Strategy, Education Investment Fund, and the Australia India Strategic Research Fund, and Astronomy Australia Limited, under contract to Curtin University. We acknowledge the iVEC Petabyte Data Store, the Initiative in Innovative Computing and the CUDA Center for Excellence sponsored by NVIDIA at Harvard University, and the International Centre for Radio Astronomy Research (ICRAR), a Joint Venture of Curtin University and The University of Western Australia, funded by the Western Australian State government.
